### 2011-11-07 doc/pldi2012/lowlevel.tex
removing fluff

The deferred nature of a deferred lock also allows a distributed implementation
of locks.  Similar to events, a deferred lock can be created on any node
without communication with the other nodes.  That node becomes
the initial home for the lock, but the deferred lock (unlike events) can 
migrate to other nodes or even be shared between nodes, allowing repeated
lock requests from a single node to be kept local to that node after a single
message is sent to the current owner.  
%(Should the ownership of the lock change
%while the message is in flight, the previous owner forwards the request to
%the new owner.  The highly unlikely, but possible, case in which multiple 
%forwardings are required is bounded at the number of nodes in the machine.)
The ability to migrate locks between nodes allows the performance of a
heavily-contended cluster-level lock to approach that of a single-node lock.

The final interesting feature of the deferred lock is that it can be associated
with an object in memory and that object will automatically be kept coherent
across the cluster.  This is similar to MESI coherence in a distributed cache,
but is done per-object rather than on cache lines.  This is currently only 
used for object within the runtime itself, but we are exploring ways to expose
it at the Legion application level.

\subsubsection{Processors}
\label{subsec:processors}
Hardware units capable of performing computation are captured as generic
processors in the low-level runtime.  Like other runtime objects, a processor
has a globally unique name and can be referred to anywhere in the system.
Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
but they all support the same
interface and execution model.  A request to spawn a task on a given processor
may come from anywhere in the system.  The request names which task should
be spawned and provides input arguments and zero or more events.  Once those
events have occurred, the task is placed into a queue of ready tasks, which
the processor works through in the order in which they were enqueued.  A
processor will generally run a task to completion before taking another, but
will try to run another task (resources permitting) if the current task needs
to wait for an event to occur.  Once a task is placed on a processor's queue,
it may not be moved or deleted.  Load-balancing of tasks is expected to be done
by the high-level runtime.  The primary purpose of the queue for the low-level
Processor is to be able to hide the latency of starting new tasks.

\subsubsection{Memories}
\label{subsec:memories}
Application-visible storage in the low-level runtime is described in terms of
memory objects.  A memory object has a size and describes its affinity to 
processors (and other memories) in terms of relative bandwidth and latency.
The low-level runtime allows for the allocation and freeing of region instances
and allocators.  It also supports bulk transfers from one memory to another.
These bulk transfers can often be implemented using dedicated DMA (or RDMA)
hardware, freeing up the processors to perform more interesting
computation.

\subsubsection{Regions}
\label{subsec:regionmeta}
Application-level regions are broken into four inter-related pieces in the
low-level runtime:
\begin{description}

\item[MetaData] The metadata for a logical region captures attributes of a
logical region that are common across all instances.  These attributes include
the size of the logical region, the element type, and the allocation mask,
which tracks which portions of the logical region are actually in use.  
%Region
%MetaDatas are created independently of any memory objects and are accessible
%(via a globally unique name) to all nodes in the machine.

\item[Allocator] The dynamic memory allocation capability of a region instance
is provided via an allocator object.  An allocator is associated with a
region metadata and must be created in a particular memory.  Once created, it
may be used to allocate and free elements of the region by any task on any
processor that has non-zero affinity with the memory.

\item[Instance] Each physical instance of a region stores some version of all
the elements in a region.  As with allocators, a physical instance is
associated with a particular region metadata and spends its entire lifetime in
the memory object in which it was created.  
%When multiple physical instances
%exist for the same region, the low-level runtime does nothing to keep them
%copies and set up the right inter-operation dependencies to provide the level
%of coherence requested by the application.  Although a physical instance 
%stores the contents of a region, it does not directly offer access to those
%contents.

\item[Accessor] Access to the contents of a physical region instance are 
handled by accessor objects.  These are lightweight objects that allow a given
task to access a given physical region instance, regardless of the
implementation of that instance or the underlying memory.  In the most common
case, read and write access is done through a single array dereference, but the
accessor interface allows seamless use of other types of memory as well (e.g.
memory that's accessed via RMDA put and get calls).

\end{description}

\subsection{Low-Level Implementations}
\label{subsec:lowimpl}
The primary goal of the low-level runtime is to provide an
abstraction that enables efficient execution of Legion programs
on a wide array of machines.  To support this claim we describe
two implementation of the low-level runtime: one based on POSIX
threads for a shared memory SMP and one based on GASNet and CUDA
for distributed memory clusters with multiple GPUs.
%The magic of an
%apparently-unified address space is left to the high-level runtime, and the
%interface to the low-level runtime is done in terms of handles rather than 
%pointers for runtime objects and relative offsets rather than absolute 
%pointers for resource allocations.

@@

\subsubsection{GPU Cluster Implementation}
\label{subsec:clusterimpl}
The cluster implementation of the low-level runtime is implemented on a mix
of POSIX threads for intra-node threading and communication, GASnet for 
inter-node communication, and CUDA for access to resources of the GPUs in 
each node.  Locks and events have roughly the same implementation as the SMP
runtime for handling multiple local requestors, but use GASnet active messages
to coordinate lock ownership and event subscription and triggering between
nodes.  In addition to the processor objects for each CPU core, a single
processor object is created for each GPU in the system.  It would be preferable
to expose the GPU's computational resources at a finer granularity, but
current GPU hardware doesn't directly allow this flexibility and recent attempts
at implementing such control in a software layer on top of CUDA have suffered
from significant overhead.  The memory hierarchy is expanded to include a 
{\em global} memory that is striped across all nodes and accessible via GASnet
RDMA calls, one process local memory per cluster node, and two memories per GPU: one for the GPU's framebuffer that is
accessible only to that GPU and by bulk DMA operations, and one for the pinned
{\em zero copy} system memory that is accessible to the GPU and all the CPU cores
in a single system.

>>>>

The deferred nature of a deferred lock also allows a distributed implementation
of locks.  Similar to events, a deferred lock can be created on any node
without communication with the other nodes.  That node becomes
the initial home for the lock, but the deferred lock (unlike events) can 
migrate to other nodes or even be shared between nodes, allowing repeated
lock requests from a single node to be kept local to that node after a single
message is sent to the current owner.  
%(Should the ownership of the lock change
%while the message is in flight, the previous owner forwards the request to
%the new owner.  The highly unlikely, but possible, case in which multiple 
%forwardings are required is bounded at the number of nodes in the machine.)
%The ability to migrate locks between nodes 
This allows the performance of a
heavily-contended cluster-level lock to approach that of a single-node lock.

The final interesting feature of the deferred lock is that it can be associated
with an object in memory and that object will automatically be kept coherent
across the cluster.  This is similar to MESI coherence in a distributed cache,
but is done per-object rather than on cache lines.  This is currently only 
used for objects within the runtime itself, but we are exploring ways to expose
it at the Legion application level.

\subsubsection{Regions}
\label{subsec:regionmeta}
Application-level regions are broken into three inter-related pieces in the
low-level runtime.  The metadata object tracks global information such as
region size and valid mask.  Allocators are used to dynamically allocate and
free entries in a region.  Finally, instances hold versions of the data for 
a logical region.

\subsubsection{Memories}
\label{subsec:memories}
Application-visible storage in the low-level runtime is described in terms of
memory objects.  A memory object has a size and describes its affinity to 
processors (and other memories) in terms of relative bandwidth and latency.
The low-level runtime allows for the allocation and freeing of region instances
and allocators.  It also supports bulk transfers from one memory to another.
These bulk transfers can often be implemented using dedicated DMA (or RDMA)
hardware, freeing up the processors to perform more interesting
computation.  Finally, the runtime is able to provide memory-specific accessors,
lightweight objects that allow a given task to access a given physical region
instance.  In the most common case, reads and write access is done through a 
single array dereference, but the interface allows seamless use of other
access modes (e.g. use of RDMA puts and gets for access to remote memory).

\subsubsection{Processors}
\label{subsec:processors}
Hardware units capable of performing computation are captured as generic
processors in the low-level runtime.  Like other runtime objects, a processor
has a globally unique name and can be referred to anywhere in the system.
Processors come in different \emph{kinds} (e.g. X86 CPU core, CUDA-capable GPU),
but they all support the same
interface and execution model.  A request to spawn a task on a given processor
may come from anywhere in the system, and names a task along with zero or
more events that must happen before the task may start.  Once those
events have occurred, the task is placed into a queue of ready tasks, which
the processor works through in the order in which they were enqueued.  A
processor will generally run a task to completion before taking another, but
will try to run another task (resources permitting) if the current task needs
to wait for an event to occur.  Once a task is placed on a processor's queue,
it may not be moved or deleted.  Load-balancing of tasks is expected to be done
by the high-level runtime.  The primary purpose of the queue for the low-level
Processor is to be able to hide the latency of starting new tasks.

\subsection{Low-Level Implementations}
\label{subsec:lowimpl}
The primary goal of the low-level runtime is to provide an
abstraction that enables efficient execution of Legion programs
on a wide array of machines.  To support this claim we describe
two implementation of the low-level runtime: one based on POSIX
threads for a shared memory SMP and one based on GASNet \cite{GASNET07} 
and CUDA for distributed memory clusters with multiple GPUs.
%The magic of an
%apparently-unified address space is left to the high-level runtime, and the
%interface to the low-level runtime is done in terms of handles rather than 
%pointers for runtime objects and relative offsets rather than absolute 
%pointers for resource allocations.

@@

\subsubsection{GPU Cluster Implementation}
\label{subsec:clusterimpl}
The cluster implementation of the low-level runtime is implemented on a mix
of POSIX threads for intra-node threading and communication, GASNet for 
inter-node communication, and CUDA for access to resources of the GPUs in 
each node.  Locks and events have roughly the same implementation as the SMP
runtime for handling multiple local requestors, but use GASNet active messages
to coordinate lock ownership and event subscription and triggering between
nodes.  In addition to the processor objects for each CPU core, a single
processor object is created for each GPU in the system.  It would be preferable
to expose the GPU's computational resources at a finer granularity, but
current GPU hardware doesn't directly allow this flexibility and recent attempts
at implementing such control in a software layer on top of CUDA have suffered
from significant overhead.  The memory hierarchy is expanded to include a 
{\em global} memory that is striped across all nodes and accessible via GASNet
RDMA calls, one process local memory per cluster node, and two memories per GPU: one for the GPU's framebuffer that is
accessible only to that GPU and by bulk DMA operations, and one for the pinned
{\em zero copy} system memory that is accessible to the GPU and all the CPU cores
in a single system.
